Explore and Summarize Data by Yongnan Sun
========================================================
[Download data](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1504549612426000&usg=AFQjCNHlCZXexmiYS4VWtm1qqupA4JiUsQ)

# Introduction

This tidy data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent).

This dataset is public available for research. The details are described in 
[Cortez et al., 2009]. 

Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016

# Summary wine data

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(dplyr)
library(tidyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
wine$X <- NULL
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(wine)

# Create a new factor variable quality.level
wine$quality.level <- ifelse(wine$quality < 5, "low", 
                             ifelse(wine$quality < 7, "average", "high"))
wine$quality.level <- factor(wine$quality.level, 
                             levels=c("high", "average", "low"), ordered=TRUE)
```

Our dataset consists of 12 variables, with 1599 observations. Quality variable 
is discrete and the others are continuous.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
qplot(as.factor(quality), data=wine, geom="bar") +
  xlab("quality")
summary(wine$quality)
```

Red wine quality is normally distributed and concentrated around 5 and 6.

```{r}
qplot(as.factor(quality.level), data=wine, geom = "bar") +
  xlab("quality.level")
summary(wine$quality.level)
```
Explore the new variable we created.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(fixed.acidity, data = wine, binwidth=0.1)
```

The distribution of fixed acidity is right skewed, and concentrated around 7.9.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(volatile.acidity, data = wine, binwidth = 0.01)
```

The distribution of volatile acidity seems to be unclear whether it is bimodal 
or unimodel, right skewed or normal.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(citric.acid, data = wine, binwidth = 0.01)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(residual.sugar, data = wine, binwidth = 0.1)
```

```{r}
qplot(residual.sugar, data = wine, binwidth = 0.05) +
  scale_x_log10(breaks = seq(0.1, 12, 2), limits = c(0.1, 12))
```

The distribution of residual sugar is right skewed, and concentrated around 2. 
There are a few outliers in the plot. Using scale_x_log10() transformer we see 
the distribution is close to normal. It can be considered a lognormal 
distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(chlorides, data = wine, binwidth = 0.01)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(chlorides, data = wine, binwidth = 0.01) +
  scale_x_continuous(limits = c(0, quantile(wine$chlorides, 0.98)))
```

If we remove the top 2 percent chlorides(outlies) we can see the distribution is nearly a
normal, and concentrated around 0.08. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(free.sulfur.dioxide, data = wine, binwidth = 1)
```

```{r}
qplot(free.sulfur.dioxide, data = wine, binwidth = 1) +
  scale_x_continuous(limits = c(0, quantile(wine$free.sulfur.dioxide, 0.99)))
```

The distribution of free sulfur dioxide is right skewed and concentrated around 
14. The second plot removed top 1 percent of outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(total.sulfur.dioxide, data = wine, binwidth = 5)
```

```{r}
qplot(total.sulfur.dioxide, data = wine, binwidth = 5) +
  scale_x_continuous(limits = c(0, quantile(wine$total.sulfur.dioxide, 0.99)))
```

The distribution of total sulfur dioxide is right skewed and concentrated around
38. The second plot removed top 1 percent of outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(density, data = wine, binwidth = 0.001)
```

The distribution of density is normal and concentrated around 0.9967.

```{r, message=FALSE, warning=FALSE}
qplot(pH, data = wine, binwidth = 0.01)
```

The distribution of pH is normal and concentrated around 3.310.

```{r, message=FALSE, warning=FALSE}
qplot(sulphates, data = wine, binwidth = 0.01)
```

```{r}
qplot(sulphates, data = wine, binwidth = 0.01) +
  scale_x_continuous(limits = c(0, quantile(wine$sulphates, 0.99)))
```

The distribution of sulphates is right skewed and concentrated around 0.6581. 
The second plot removed top 1 percent of outliers.

```{r, message=FALSE, warning=FALSE}
qplot(alcohol, data = wine, binwidth = 0.5)
```

The distribution of alcohol is right skewed and concentrated around 10.20.


# Univariate Analysis


### What is the structure of your dataset?
There are 1,599 red wines in the dataset with 11 features on the chemical 
properties of the wine.

The median quality is 6. Most wines have a pH of 3.4 or higher. About 75% of 
wine have quality that is lower than 6. The median percent alcohol content is 
10.20 and the max percent alcohol content is 14.90.


### What is/are the main feature(s) of interest in your dataset?
The main features in the data set are pH and quality. I'd like to determine 
which features are best for predicting the quality of a wine. I suspect pH and 
some combination of the other variables can be used to build a predictive model 
to grade the quality of wines.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Volatile acidity, citric acid, and alcohol likely contribute to the quality of a
wine. I think volatile acidity (the amount of acetic acid in wine) and alcohol 
(the percent alcohol content of the wine) probably contribute most to the 
quality after researching information on wine quality.


### Did you create any new variables from existing variables in the dataset?
I created a new variable called "quality.level" which is categorically divided 
into "low", "average", and "high". This grouping method will help us detect the 
difference among each group more easily.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Having visualized acitric acid and volatile acidity data, I observed some 
unusual distributions, so I guess this fact may have some correlation with the 
quality of red wine. Since the data is clean, I did not perform any cleaning 
process or modification of the data.


# Bivariate Plots Section

```{r}
# Select subset of the data to save time
addFeatures <- wine[,!colnames(wine) %in% c("volatile.acidity", 
                                            "quality", "quality.level")]
ggpairs(addFeatures, 
        columnLabels=c("f.aci", "ci.aci", "res.sug", "chlo", "fr.su.dio", 
                       "to.su.dio", "dens", "pH", "sulph", "alco"), 
        lower = list(continuous = wrap("points", size=1, shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.')))) + 
  theme(axis.ticks=element_blank(),
        axis.line=element_blank(), 
        axis.text=element_blank(), 
        panel.grid.major= element_blank())
```

This correlation plot shows a positive relationship between density and fixed acidity, a 
positive relationship between fixed acidity and citric acid, and a negative 
relationship between pH and acidity. After this we can analyze the correlations 
between variables in more details.

```{r echo=FALSE, message=FALSE, warning=FALSE}
attach(wine)
# Create a function for bivariate plots with 3 inputs
bi_qplot <- function(x, y, z="jitter") {
  if(z=="boxplot") {
    qplot(x=x, y=y, data=wine, geom="jitter", alpha=0.01) + 
      geom_boxplot() +
      guides(alpha="none") +
      theme_bw()
  }
  else {
    qplot(x=x, y=y, data=wine, geom=z, alpha=0.01) + 
      guides(alpha="none") +
      theme_bw()
  }
}


bi_qplot(quality.level, volatile.acidity, "boxplot") +
  xlab("Quality level") +
  ylab("Volatile acidity (g/dm^3)")

cor(wine$quality, wine$volatile.acidity)
```

The graph shows a very clear trend that the lower volatile acidity is, the 
higher the quality becomes. The correlation coefficient between quality and 
volatile acidity is -0.39. This can be explained by the fact that volatile 
acidity at too high of levels can lead to an unpleasant, vinegar taste.

```{r, message=FALSE, warning=FALSE}
bi_qplot(quality.level, citric.acid, "boxplot") +
  xlab("Quality level") +
  ylab("Citric acid")

cor(wine$quality, wine$citric.acid)
```

The correlation coefficient is 0.226. The graph shows a weak positive 
relationship between quality level and citric acid.

```{r, message=FALSE, warning=FALSE}
bi_qplot(quality.level, alcohol, "boxplot") +
  xlab("Quality level") +
  ylab("Alcohol")

cor(wine$quality, wine$alcohol)
```

With the correlation coefficient of 0.476, the graph shows a positive 
relationship between alcohol and quality level.


```{r, message=FALSE, warning=FALSE}
bi_qplot(residual.sugar, alcohol) +
  xlab("Residual sugar (g/dm^3)") +
  ylab("Alcohol (%)") +
  scale_x_continuous(limits = c(0, quantile(wine$residual.sugar, 0.99)))
```

If we actually examine winemaking process, we see that there is a global trend 
for wines that are made from ripe to overly ripe grape fruit. To keep wines from
staying too sweet, the fermentation process has to be left to continue until 
more of the sugar is consumed, but as a byproduct, more alcohol is present in 
the wines.

```{r, message=FALSE, warning=FALSE}
bi_qplot(citric.acid, volatile.acidity) +
  xlab("Citric acid (g/dm^3)") +
  ylab("Volatile acidity (g/dm^3)")

cor(wine$citric.acid, wine$volatile.acidity)
```

There is a negative correlation between citric acid and volatile acidity.

```{r, message=FALSE, warning=FALSE}
bi_qplot(alcohol, density) + 
  xlab("Alcohol (%)") + 
  ylab("Density (g/cm^3)")

cor(wine$alcohol, wine$density)
```

The correlation coefficient is -0.50, so the relationship is quite clear. As 
percent of alcohol content increases, the density decreases. The reason is that 
the density of wine is lower than the density of pure water.

```{r}
# Create a function for density plots. First input is x, second input for 
# selecting color, third input for writing label name
den_plot <- function(x, color, xlab) {
  ggplot(data=wine, aes(x, color=color)) + 
    geom_density() + 
    xlab(xlab) + 
    labs(color="Quality level") +
    theme_bw()
}
den_plot(fixed.acidity, quality.level, "Fixed acidity (g/dm^3)")
```

The distribution of low and average quality wines seem to be concentrated at 
fixed acidity values between 6 and 10. pH increases as fixed acidity 
decreases, and citric acid increases as fixed acidity increases.

```{r}
den_plot(alcohol, quality.level, "Alcohol (%)")
```

High quality wine density line is distinct from the others, and mostly 
distributed between 11 and 12.

```{r}
den_plot(volatile.acidity, quality.level, "Volatile acidity (g/dm^3)")
```

This chart shows a very clear trend; as volatile acidity decreases, the quality 
of wine increases. Wines with volatile acidity exceeding 1 are almost rated as 
low quality.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
I observed a negative relationship between quality level and volatile acidity, 
and a positive relationship between quality level and alcohol. I am not suprised 
at this result because men tend to grade stronger wines as high quality, whereas 
wines with low percent of alcohol are often not graded as such. High volatile 
acidity is also perceived to be undesirable because it impacts the taste of 
wines, which at too high of levels can lead to an unpleasant, vinegar taste 
according to the data information.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
There is a positive relationship between density and fixed acidity, a positive 
relationship between fixed acidity and citric acid, and a negative relationship 
between pH and fixed acidity.

### What was the strongest relationship you found?
Quality is positively and strongly correlated with alcohol, and it is also 
negatively correlated with volatile acidity. Alcohol and volatile acidity could 
be used in a model to predict the quality of wine.

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.level, 
           size = citric.acid), data = wine) +
  geom_point(position = position_jitter(h = 0), alpha = 0.3) +
  scale_x_continuous(limits = c(0, quantile(wine$volatile.acidity, 0.99)))

```

High quality feature seems to be associated with alcohol ranging from 11 to 13, 
volatile acidity from 0.2 to 0.5, and citric acid from 0.25 to 0.75

```{r message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = fixed.acidity, color = quality.level, 
           size = pH), data = wine) +
  geom_point(position = position_jitter(w = 0.025, h = 0.025), alpha = 0.3) +
  scale_x_continuous(limits = c(0, quantile(wine$volatile.acidity, 0.99)))
  
```

From this plot we can also see that high quality is associated with low volatile 
acidity and reletively high pH. However, it is spread out in the direction of 
fixed acidity.

```{r message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol, color = quality.level), 
       data = wine) +
  geom_point(position = position_jitter(w = 0.025, h = 0.025), alpha = 0.3) +
  scale_x_continuous(limits = c(0, quantile(wine$residual.sugar, 0.99)))
```

Residual sugar has no great effect on quality level.

```{r message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = alcohol, color = quality.level, 
           size = citric.acid), data = wine) +
  geom_point(position = position_jitter(w = 0.025, h = 0.025), alpha = 0.3)
```

## Try some linear models

```{r message=FALSE, warning=FALSE}
alcohol_lm <- lm(data=wine, quality~alcohol)
summary(alcohol_lm)
```

```{r message=FALSE, warning=FALSE}
volacid_lm <- lm(data=wine, quality~volatile.acidity)
summary(volacid_lm)
```

```{r message=FALSE, warning=FALSE}
feaInterest_lm <- lm(data=wine, quality~volatile.acidity + alcohol)
summary(feaInterest_lm)
```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
When looking at wine quality level, we see a positive relationship between fixed
acidity and citric acid.

### Were there any interesting or surprising interactions between features?
Residual sugar, supposed to play an important part in wine taste, actually has 
very little impact on wine quality.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
es, I created 3 models. Their p-values are significant; however, the R squared 
values are under 0.4, so they do not provide us with enough explanation about 
the variability of the response data around their means.
------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data=wine, aes(factor(quality), fill=quality.level)) + 
  geom_bar() + 
  xlab("Quality") + 
  ylab("Number of wines") +
  ggtitle("Number of wines in different qualities")
```

### Description One
The distribution of red wine quality appears to be normal. 82.5% of wines are 
rated 5 and 6 (average quality). Although the rating scale is between 0 and 10, 
there exists no wine that is rated 1, 2, 9 or 10.

### Plot Two
```{r echo=FALSE, Plot_Two}
bi_qplot(quality.level, citric.acid, "boxplot") +
  xlab("Quality level") +
  ylab("Citric acid (g/dm^3)") +
  ggtitle("Citric acid level")
```

### Description Two
While citric citric do not have a strong correlation with quality, it is an 
important component in the quality of wine. Because citric acid is an organic 
acid that contributes to the total acidity of a wine, it is crucial to have a 
right amount of citric acid in wine. Adding citric acid will give the wine 
"freshness" otherwise not present and will effectively make a wine more acidic. 
Wines with citric acid exceeding 0.75 are hardly rated as high quality. 50% of 
high quality wines have a relatively high citric acid that ranges between 0.3 
and 0.49, whereas average and low quality wines have lower amount of citric 
acid.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.level, 
           size = citric.acid), data = wine) +
  geom_point(position = position_jitter(h = 0), alpha = 0.3) +
  ggtitle("Quality level with alcohol, volatile acidity and citric acid") +
  xlab("Volatile acidity (g/dm^3)")
```

### Description Three
People tend to grade stronger wines as high quality, whereas wines with low 
percent of alcohol are often not graded as such. High volatile acidity is also 
perceived to be undesirable because it impacts the taste of wines, which at too 
high of levels can lead to an unpleasant, vinegar taste. Citric acid can add 
'freshness' and flavor to wines.

Wine with high percent alcohol content and low volatile acidity tends to be 
rated as high quality wine. Based on the result, we can see that the volatile 
acidity in wine and percent alcohol content are two important components in the 
quality and taste of red wines.
------

# Reflection


The wines data set contains information on 1599 wines across twelve variables. 
I started by understanding the individual variables in the data set, and then I 
explored interesting questions and make observations on plots. Eventually, I 
explored the quality of wines across many variables and tried to create 3 linear 
models to predict red wine quality.

Citric acidity and residual sugar which seem to me have a great impact on taste 
of wine actually do not reflect much of the quality. During the exploration, I 
found it is hard to find the distribution between quality and other features, 
since quality is not a continuous variable.

There are very few wines that are rated as low or high quality. We could improve
the quality of our analysis by collecting more data, and creating more variables
that may contribute to the quality of wine. This will certainly improve the 
accuracy of the prediction models. Having said that, we have successfully 
identified features that impact the quality of red wine, visualized their 
relationships and summarized their statistics.